Supplementary Information

### 2021-06-01 doc/manuscript/ForestGEO_treerings.Rmd
cut recruitment plots & climwin plot references from manuscript
#130

#SI figure numbers
all_recruitment_plots="S1-S10"

all_traditional_comparisons = "S11-S14"
SCBI_traditional_comparison = "S11"
Zofin_traditional_comparison = "S12"
CB_traditional_comparison = "S13"
SC_traditional_comparison = "S14"

all_climwin = "S15-S34"
climwin_BCNM_W = "S15"
climwin_BCNM_T = "S16"
climwin_HKK_W = "S17"
climwin_HKK_T = "S18"
climwin_SCBI_W = "S19"
climwin_SCBI_T = "S20"
climwin_HF_W = "S23"
climwin_HF_T = "S24"

@@

We analyzed tree-ring data, most most of which was collected for earlier studies (see references in Table `r sites`), from 10 sites ranging from 9.15$^\circ$ to 61.30$^\circ$ N latitude and representing a wide range of forest and tree types: tropical broadleaf deciduous and evergreen, temperate broadleaf deciduous and conifer, and boreal conifer (Tables `r sites`, `r site_details`, `r species_table`).
Nine of these sites (exception: LT) were co-located with large forest dynamics plots of the Forest Global Earth Observatory [ForestGEO, @anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021].
Trees were cored within the ForestGEO plots (n=5 sites) and/or nearby within similar forest types (n=5 sites), following a variety of sampling protocols designed to meet the varied objectives of the original studies (Tables `r site_details`, `r sampling_details_table`). 
There was wide variation in the distribution of record start years, which approximate the age distribution of sampled trees (Figs. `r all_recruitment_plots`).
In using this diversity of data sources, we ensured that our approach could handle challenges presented by varying methodologies and forest types.

@@

Data analysis consisted of two main steps: (1) identifying the primary climate drivers (i.e., variables and seasonal windows over which they are most influential on tree growth), and (2) combining these climate drivers, DBH, and year into a multivariate model (Fig. `r schematic`). 
The analysis was run separately for each site (step 1), site-species combination (step 2), and each response variable estimating different measures of tree growth (RW, BAI, or $\Delta$AGB). 
We note that the decision to identify primary climate drivers at the level of site, as opposed to species, was motivated by the expectation that differences in the most influential climate drivers across species in one site would be minimal compared to cross-site differences (Fig. `r SCBI_example`); however, analyses focused on interspecific differences could optimize species-specific climate sensitivity estimates by fitting *climwin* individually by species.

@@

We used the *climwin* package in R [@vandepol_identifying_2016] to identify the most important climate variable and the seasonal window over which its effect was strongest for each of two categories of variables: a temperature group ($T_{min}$, $T_{max}$, $T_{mean}$, and PET) and a precipitation group (PPT, PDF).
To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by fitting penalized thin-plate regression splines in generalized additive models [GAM, functions *gam* and *s* in the R package *mgcv*, @wood_fast_2011] to individual growth records (RW, BAI, or $\Delta$AGB) from each core, and extracted the residual variation for each observation. 
The smoothing parameters were automatically selected by the *gam* function with generalized cross-validation (GCV).
We then used *climwin* to identify the climate variables that most strongly correlated with the individual tree-level residuals of the growth variables, RW, BAI, or $\Delta$AGB, specifying quadratic fits to allow for potential nonlinearities in the climate response.
Within *climwin*, we specified a mixed-effects model in which the fixed effects were the climate variables and the random intercepts were species (when n $\ge$ 3) and core identity (noting that these effects should be minimal given that residuals are centered around zero).
For each climate variable, we ran permutations for all possible combinations of consecutive months within a 15-month period ending near the time of cessation of formation of each annual ring (Table `r site_details`). 
*Climwin* runs all potential models to select the best fit (lowest AIC), and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@vandepol_identifying_2016]. 
For each group of candidate climate variables (water and temperature; Fig. `r schematic`), we selected the variable - seasonal window combination identified by *climwin* as having the lowest AIC as a candidate climate variable for the multivariate models.

@@

### Most influencial climate drivers {-#Most}
```{r , eval=FALSE, echo=FALSE}
## Tallies for the following paragraph (Feb 11)
#sameV_sameW: LDWp, NMt,NMp, NIOt, NIOp, SCBIp,  ZOFp,ZOFt,  (8)
#sameV_diffW:      (10)
  # overlapping: BCNMp, BCNMt, HKKp, SCBIt, LDWt, CBp,CBt , SCp, SCt, (9)
  # non-overlapping: HFt, (1)
#diffV_diffW: HFp (non-overlapping), HKKt (overlapping)
```
At each site, the three metrics of growth (RW, BAI, and $\Delta$AGB) exhibited similar patterns in the direction of response and relative strength of correlation to climate variables across the range of potential seasonal windows. 
However, the seasonal window exhibiting the strongest climatic effect on growth and even the most influential climate variable sometimes differed among the growth metrics (Figs. `r all_climwin`).
For eight of 20 site-variable group (i.e., water and temperature) combinations, both the most influential climate variable and seasonal window were identical across growth metrics (e.g., Fig. `r climwin_SCBI_W`). 
For nine site-variable group combinations, *climwin* identified the same climate variable and overlapping seasonal windows  (e.g., Fig. `r climwin_SCBI_T`), 
and in one case (at HKK) different variables ($T_{max}$ and $T_{mean}$) were selected with overlapping seasonal windows (Fig. `r climwin_HKK_T`). 
For just two site-variable group combinations (both variable groups at HF, where climate had only weak effects and mixed responses among species in the final models), *climwin* identified completely different seasonal windows and, for precipitation, different variables (PPT and PDF; Figs. `r climwin_HF_W`- `r climwin_HF_T`).
Henceforth, unless otherwise noted, we focus on the climate sensitivities identified using RW as the growth metric and for the full set of cores (i.e., including those for which DBH could not be reconstructed).

@@

In all seven sites dominated by trees less than 100 years old, growth trends were universally negative, when significant (Table `r hypotheses`, Fig. `r year_fig`). 
These trends are probably attributable primarily to stand dynamics as cohorts and stands develop over time, and are fairly typical of mixed-species stands that experience vertical stratification [@oliver_forest_1990]. 
For species exhibiting a pulse of recruitment in the past followed by little subsequent recruitment (e.g., *Acer rubrum* and *Betula alleghaniensis* at HF; Figs. `r all_recruitment_plots`), persistent differences in growth rates among individuals could produce a trend of declining growth, as faster-growing individuals reach various size thresholds earlier [@brienen_tree_2017; see also @vandersleen_trends_2017].
Particularly in secondary stands where many of the sampled species recruited in pulses that were followed by low recruitment (e.g., SCBI, HF; Appendix `r site_details_appendix`), growth declines are consistent with the tendency for faster tree growth during early stand development [@lorimer_estimating_1988; @lorimer_methodology_1989; @oliver_forest_1990], and with increasing competition and declining woody productivity as young stands mature [e.g., @goulden_patterns_2011; @pregitzer_carbon_2004].
Abiotic drivers likely also play a role at some of these sites.
At Scotty Creek, in northern Canada, rapid warming is thawing permafrost and altering hydrologic conditions [@baltzer_forests_2014], resulting in high mortality, growth declines, and low recruitment of *Picea mariana* [@sniderhan_growth_2016; @dearborn_permafrost_2020]; we attribute pronounced negative growth trends to a combination of successional declines and indirect climatic stress.

@@

Figures `r all_recruitment_plots`. Density plots of core record start years by species for all sites.

@@

Figures `r all_climwin`. Climwin output for water and temperature variable groups at all sites.

>>>>

#SI figure numbers
##traditional comparisons
all_traditional_comparisons = "S1-S4"
SCBI_traditional_comparison = "S1"
Zofin_traditional_comparison = "S2"
CB_traditional_comparison = "S3"
SC_traditional_comparison = "S4"

##climwin
climwin_HKK_T = "S18"
climwin_SCBI_W = "S19"
climwin_SCBI_T = "S20"
climwin_HF_W = "S23"

@@

We analyzed tree-ring data, most most of which was collected for earlier studies (see references in Table `r sites`), from 10 sites ranging from 9.15$^\circ$ to 61.30$^\circ$ N latitude and representing a wide range of forest and tree types: tropical broadleaf deciduous and evergreen, temperate broadleaf deciduous and conifer, and boreal conifer (Tables `r sites`, `r site_details`, `r species_table`).
Nine of these sites (exception: LT) were co-located with large forest dynamics plots of the Forest Global Earth Observatory [ForestGEO, @anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021].
Trees were cored within the ForestGEO plots (n=5 sites) and/or nearby within similar forest types (n=5 sites), following a variety of sampling protocols designed to meet the varied objectives of the original studies (Tables `r site_details`, `r sampling_details_table`). 
In using this diversity of data sources, we ensured that our approach could handle challenges presented by varying methodologies and forest types.

@@

Data analysis consisted of two main steps: (1) identifying the primary climate drivers (i.e., variables and seasonal windows over which they are most influential on tree growth), and (2) combining these climate drivers, DBH, and year into a multivariate model (Fig. `r schematic`). 
The analysis was run separately for each site (step 1), site-species combination (step 2), and each response variable estimating different measures of tree growth (RW, BAI, or $\Delta$AGB). 
We note that the decision to identify primary climate drivers at the level of site, as opposed to species, was motivated by the expectation that differences in the most influential climate drivers across species in one site would be minimal compared to cross-site differences (Fig. `r SCBI_example`); however, analyses focused on interspecific differences could optimize species-specific climate sensitivity estimates by fitting individually by species.

@@

We used the *climwin* package in R [@vandepol_identifying_2016] to identify the most important climate variable and the seasonal window over which its effect was strongest for each of two categories of variables: a temperature group ($T_{min}$, $T_{max}$, $T_{mean}$, and PET) and a precipitation group (PPT, PDF).
To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by fitting penalized thin-plate regression splines in generalized additive models [GAM, functions *gam* and *s* in the R package *mgcv*, @wood_fast_2011] to individual growth records (RW, BAI, or $\Delta$AGB) from each core, and extracted the residual variation for each observation. 
The smoothing parameters were automatically selected by the *gam* function with generalized cross-validation (GCV).
We then used *climwin* to identify the climate variables that most strongly correlated with the individual tree-level residuals of the growth variables, RW, BAI, or $\Delta$AGB, specifying quadratic fits to allow for potential nonlinearities in the climate response.
Within *climwin*, we specified a mixed-effects model in which the fixed effects were the climate variables and the random intercepts were species (when n $\ge$ 3) and core identity (noting that these effects should be minimal given that residuals are centered around zero).
For each climate variable, we ran permutations for all possible combinations of consecutive months within a 15-month period ending near the time of cessation of formation of each annual ring (Table `r site_details`). 
*Climwin* runs all potential models to select the best fit (lowest AIC), and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@vandepol_identifying_2016]. 
For each group of candidate climate variables (water and temperature; Fig. `r schematic`), we selected the variable - seasonal window combination with the lowest AIC as a candidate climate variable for the multivariate models.

@@

### Most influencial climate drivers {-#Most}
```{r , eval=FALSE, echo=FALSE}
## Tallies for the following paragraph (Feb 11)
#sameV_sameW: LDWp, NMt,NMp, NIOt, NIOp, SCBIp,  ZOFp,ZOFt,  (8)
#sameV_diffW:      (10)
  # overlapping: BCNMp, BCNMt, HKKp, SCBIt, LDWt, CBp,CBt , SCp, SCt, (9)
  # non-overlapping: HFt, (1)
#diffV_diffW: HFp (non-overlapping), HKKt (overlapping)

```
At each site, the three metrics of growth (RW, BAI, and $\Delta$AGB) exhibited similar patterns in the direction of response and relative strength of correlation to climate variables across the range of potential seasonal windows. 
However, the seasonal window exhibiting the strongest climatic effect on growth and even the most influential climate variable sometimes differed among the growth metrics.
For eight of 20 site-variable group (i.e., water and temperature) combinations, both the most influential climate variable and seasonal window were identical across growth metrics (e.g., PPT at SCBI; Fig. `r climwin_SCBI_W`). 
For nine site-variable group combinations, *climwin* identified the same climate variable and overlapping seasonal windows  (e.g., PET at SCBI Fig. `r climwin_SCBI_T`), 
and in one case (at HKK) different variables ($T_{max}$ and $T_{mean}$) were selected with overlapping seasonal windows (Fig. `r climwin_HKK_T`). 
For just two site-variable group combinations (both variable groups at HF, where climate had only weak effects and mixed responses among species in the final models), *climwin* identified completely different seasonal windows and, for precipitation, different variables (PPT and PDF; Fig. `r climwin_HF_W`).
Henceforth, unless otherwise noted, we focus on the climate sensitivities identified using RW as the growth metric and for the full set of cores (i.e., including those for which DBH could not be reconstructed).

@@

In all seven sites dominated by trees less than 100 years old, growth trends were universally negative, when significant (Table `r hypotheses`, Fig. `r year_fig`). 
These trends are probably attributable primarily to stand dynamics as cohorts and stands develop over time, and are fairly typical of mixed-species stands that experience vertical stratification [@oliver_forest_1990]. 
For species exhibiting a pulse of recruitment in the past followed by little subsequent recruitment (e.g., *Acer rubrum* and *Betula alleghaniensis* at HF), persistent differences in growth rates among individuals could produce a trend of declining growth, as faster-growing individuals reach various size thresholds earlier [@brienen_tree_2017; see also @vandersleen_trends_2017].
Particularly in secondary stands where many of the sampled species recruited in pulses that were followed by low recruitment (e.g., SCBI, HF; Appendix `r site_details_appendix`), growth declines are consistent with the tendency for faster tree growth during early stand development [@lorimer_estimating_1988; @lorimer_methodology_1989; @oliver_forest_1990], and with increasing competition and declining woody productivity as young stands mature [e.g., @goulden_patterns_2011; @pregitzer_carbon_2004].
Abiotic drivers likely also play a role at some of these sites.
At Scotty Creek, in northern Canada, rapid warming is thawing permafrost and altering hydrologic conditions [@baltzer_forests_2014], resulting in high mortality, growth declines, and low recruitment of *Picea mariana* [@sniderhan_growth_2016; @dearborn_permafrost_2020]; we attribute pronounced negative growth trends to a combination of successional declines and indirect climatic stress.

@@

@@

Figure `r climwin_HKK_T_legend`

Figure `r climwin_SCBI_W_legend`

Figure `r climwin_SCBI_T_legend`

Figure `r climwin_HF_W_legend`
